---
title: "Partial Credit Model Analysis with easyRaschBayes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Partial Credit Model Analysis with easyRaschBayes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Only evaluate model-dependent chunks if the pre-fitted model exists
EVAL <- file.exists("fits/fit_pcm.rds")
```

## Overview

This vignette demonstrates a complete Partial Credit Model (PCM) Rasch analysis
workflow using `easyRaschBayes`. The analysis uses the `eRm::pcmdat2` dataset, a
small polytomous item response data set included in the `eRm` package.

All functions in this package work with a `brms` model object fitted with the
`acat` (adjacent categories) family, which parameterises the PCM.

## Data Preparation

```{r libraries, message = FALSE, warning = FALSE}
library(easyRaschBayes)
library(brms)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
```

`eRm::pcmdat2` is in wide format with item responses coded 0, 1, 2, …. The
`brms` `acat` family expects response categories starting at **1**, so we add 1
to every response before reshaping to long format.

```{r data-prep}
df_pcm <- eRm::pcmdat2 %>%
  mutate(across(everything(), ~ .x + 1)) %>%
  rownames_to_column("id") %>%
  pivot_longer(!id, names_to = "item", values_to = "response")

head(df_pcm)
```

## Fitting the PCM

The model is fitted once and saved to disk. The code chunk below shows the
fitting call (not evaluated during `R CMD check`). A pre-fitted model stored at
`fits/fit_pcm.rds` is loaded instead.

```{r define-priors}
prior_pcm <- prior("normal(0, 5)", class = "Intercept") +
  prior("normal(0, 3)", class = "sd", group = "id")
```

```{r fit-model, eval = FALSE}
fit_pcm <- brm(
  response | thres(gr = item) ~ 1 + (1 | id),
  data    = df_pcm,
  family  = acat,
  prior   = prior_pcm,
  chains  = 4,
  cores   = 4,
  iter    = 2000,
  backend = "cmdstanr"
)
saveRDS(fit_pcm, "fits/fit_pcm.rds")
```

```{r load-model, eval = EVAL}
fit_pcm <- readRDS("fits/fit_pcm.rds")
```

## Item Fit: Infit and Outfit Statistics

`infit_statistic()` computes posterior predictive infit and outfit statistics
for each item. Values near 1.0 indicate good fit; values substantially above 1
suggest underfit (unexpected responses), values below 1 suggest overfit (too
predictable).

The `ndraws_use` argument limits the number of posterior draws used, which
speeds up computation during exploration. For final reporting, use all draws
(set `ndraws_use = NULL` or omit it).

```{r infit, eval = EVAL}
fit_stats <- infit_statistic(fit_pcm, df_pcm, ndraws_use = 500)

# Posterior predictive p-values summarised per item
fit_stats %>%
  group_by(item) %>%
  summarise(
    infit_mean  = mean(infit),
    outfit_mean = mean(outfit),
    infit_ppp   = mean(infit > 1),
    outfit_ppp  = mean(outfit > 1),
    .groups = "drop"
  )
```

Posterior predictive p-values (`*_ppp`) close to 0.5 indicate that the observed
statistic falls near the centre of the posterior predictive distribution,
implying good fit. Values near 0 or 1 warrant further investigation.

## Item–Rest Score Association

`item_restscore_statistic()` computes Goodman-Kruskal's gamma between each
item's observed responses and the rest score (total score minus the focal item).
In a well-fitting Rasch model, gamma should be positive and moderate; very high
values may indicate redundancy, very low or negative values suggest the item
does not relate well to the latent trait.

```{r restscore, eval = EVAL}
rest_stats <- item_restscore_statistic(fit_pcm, df_pcm, ndraws_use = 500)

rest_stats %>%
  group_by(item) %>%
  summarise(
    gamma_mean = mean(gamma),
    gamma_ppp  = mean(gamma < 0),
    .groups = "drop"
  )
```

## Dimensionality: Residual PCA

`plot_residual_pca()` performs a principal components analysis on the
person-item residuals and plots the loadings on the first contrast. Substantial
loadings on the first contrast (eigenvalue > 1.5–2 is a common heuristic)
suggest multidimensionality.

```{r pca-plot, eval = EVAL, fig.width = 6, fig.height = 4}
plot_residual_pca(fit_pcm, df_pcm)
```

Items with positive loadings cluster on one end, negative loadings on the other.
If the contrast eigenvalue is small the residuals are essentially noise,
supporting unidimensionality.

## Local Dependence: Q3 Residual Correlations

`q3_statistic()` computes Yen's Q3 statistic — the correlation between
person-item residuals for every item pair. After conditioning on the latent
trait, residuals should be uncorrelated; elevated Q3 values indicate local
dependence (LD). A common rule of thumb is to flag pairs with Q3 above the
average Q3 plus 0.2.

```{r q3, eval = EVAL}
q3_stats <- q3_statistic(fit_pcm, df_pcm, ndraws_use = 500)

# Mean Q3 baseline (model-expected correlation under Rasch assumptions)
q3_baseline <- mean(q3_stats$q3_mean, na.rm = TRUE)

# Flag pairs exceeding the baseline by 0.2
q3_stats %>%
  filter(q3_mean > q3_baseline + 0.2) %>%
  arrange(desc(q3_mean))
```

Pairs flagged as locally dependent should be examined for substantive overlap
in item content.

## Person–Item Targeting

`plot_targeting()` produces a Wright map (person–item targeting plot) showing
the distribution of person locations alongside the item threshold locations on
the same logit scale. Good targeting occurs when person and item distributions
overlap substantially.

```{r targeting, eval = EVAL, fig.width = 6, fig.height = 5}
plot_targeting(fit_pcm, df_pcm)
```

If the person distribution is systematically above or below the item thresholds,
the test may be too easy or too hard for the sample.

## Reliability: Relative Measurement Uncertainty

`RMUreliability()` provides a Bayesian reliability estimate via Relative
Measurement Uncertainty (RMU). It requires a matrix of person location draws
with dimensions \[persons × draws\].

The `brms::ranef()` function with `summary = FALSE` returns a 3-D array
\[draws × persons × effects\]; we extract the `Intercept` slice and transpose.

```{r rmu, eval = EVAL}
person_draws <- brms::ranef(fit_pcm, summary = FALSE)$id[, , "Intercept"]
rmu <- RMUreliability(t(person_draws))
rmu
```

RMU values closer to 1 indicate that the measurement error is small relative to
the variance in person locations, implying high reliability.
